Efficient ancestry and mutation simulation with msprime 1.0

### 2021-12-01 paper.tex

\begin{abstract}
Stochastic simulation is a key tool in population genetics,
since the models involved are often analytically intractable
and simulation is usually the only way of obtaining ground-truth data to
evaluate inferences.
Because of this necessity, a large number
of specialised simulation programs have been developed, each
filling a particular niche, but with largely overlapping functionality
and a substantial duplication of effort.
Here, we introduce \msprime\ version 1.0, which
efficiently implements ancestry and mutation simulations based on
the succinct tree sequence data structure and \tskit\ library.
We summarise \msprime's many features, and show that
its performance is excellent, often many times faster
and more memory efficient than specialised alternatives.
These high-performance features have been thoroughly
tested and validated, and built using a collaborative,
open source development model,
which reduces duplication of effort and promotes
software quality via community engagement.
% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
\end{abstract}

@@

The coalescent
process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is an efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from an idealised population model, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data,
and used these simulations to characterise an analytically
intractable distribution.
This inherent efficiency, and the great utility of simulations for a wide
range of purposes, has led to dozens of different tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

Two technological developments of recent years, however,
pose major challenges to most existing simulation methods.
Firstly, fourth-generation sequencing technologies have made
complete chromosome-level assemblies possible~\citep{miga2020telomere},
% Some citations here? Darwin tree of life maybe?
and high quality assemblies are now available for many species.
Thus, modelling genetic variation data as a series of unlinked
non-recombining loci
is no longer a reasonable approximation, and
we must fully account for recombination.
However, while a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination is far more complex,
and programs such as
Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination.
The second challenge facing simulation methods is that
sample sizes in genetic studies have grown very quickly
in recent years, enabled by the precipitous fall in genome sequencing costs.
Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm~\citep{staab2015scrm} simply cannot
cope with such a large number of samples.

@@

\section*{Results}
In the following sections we describe the main features of \msprime\ 1.0,
focusing on the aspects that are either new for this version, or in which
our approach differs significantly from classical methods
(summarised in Table~\ref{tab-v1-summary}).
Where appropriate,
we benchmark \msprime\ against other simulators, but the comparisons are
illustrative and not intended to be systematic or exhaustive. Please
see~\cite{kelleher2016efficient} for a performance comparison of
\msprime\ against simulators such as \ms, \msms, and \scrm.

\begin{table}
\small{
{ % begin box to localize effect of arraystretch change
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{lp{6.5cm}}
Interface
& Separation of ancestry and mutation simulations.
Ability to store arbitrary metadata along with simulation
results, and automatic recording of provenance information
for reproducibility.
Jupyter notebook~\citep{kluyver2016jupyter} integration.
Rich suite of analytical and visualisation methods via \tskit\ library. \\

Ancestry
& SMC, SMC', Beta- and Dirac-coalescent, discrete time Wright-Fisher,
and selective sweep models. Instantaneous bottlenecks.
Discrete or continuous genomic coordinates, arbitary ploidy,
gene conversion.
Output full ARG with recombination nodes, ARG likelihood calculations.
Record full migration history and census
events.  Improved performance for large numbers of populations.
Integration with forward simulators such as \SLiM\ and \fwdpy\
(``recapitation'').
\\

@@

The majority of simulation packages are controlled either through
a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a text-based input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or a mixture of both.
Command line interfaces make it easy to run simple
simulations, but as model complexity and the number of parameters increase,
they become difficult to understand and
error-prone~\citep{ragsdale2020lessons,gower2021demes}.
Specifying parameters through a text file alleviates this problem to a degree,
but lacks flexibility, for example, when running simulations with parameters
drawn from a distribution. In practice, for any reproducible simulation
project users will write a script
to generate the required command lines or input parameter files,
invoke the simulation engine, and process the results in some way.
This process is cumbersome and labour intensive, and
a number of packages have been developed
to allow simulations to be run directly in a high-level
scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.

The more recent trend has been to move away from this file and command-line
driven approach and to instead provide direct interfaces to the simulation
engines via an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
The primary interface for \msprime\ is through a thoroughly documented
Python
API, which has encouraged the development of an ecosystem of downstream
% Found these through the Dependent packages on GitHub
tools~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
adrion2020community,adrion2020predicting, kamm2020efficiently,
mckenzie2020ipcoal, montinaro2020revisiting,
de2021geonomics,rivera2021simulation}.
As well as providing a stable and efficient platform for building
downstream applications, \msprime's Python API makes it much easier to
build reproducible simulation pipelines, as the entire workflow can
be encapsulated in a single script, and package and version
dependencies explicitly stated using the \texttt{pip}
or \texttt{conda} package managers.
For example, the errors made
in the influential simulation analysis of
\cite{martin2017human} were only detected because the pipeline
could be easily run and reanalysed~\citep{ragsdale2020lessons,martin2020erratum}.

\begin{figure*}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualisation of the separation between ancestry and mutation
        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
        is two trees along a 1kb chunk of genome relating three diploid samples.
        Each diploid individual consists of two genomes (or nodes), indicated
        by colour.
        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
        which adds mutations.
        Graphics produced using \tskit's \texttt{draw\_svg} method.
    }
\end{figure*}

@@

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its use of the ``succinct tree sequence''
data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
was introduced by \cite{kelleher2016efficient}
to concisely encode genetic ancestry and sequence variation
and was originally implemented as part of \msprime.
We subsequently extracted the core tree sequence functionality
from \msprime\ to create the \tskit\ library,
which provides a large suite of tools for processing genetic ancestry
and variation data via APIs
in the Python and C languages~\citep{tskit2021tskit}.
The availability of \tskit\ as a liberally licensed (MIT) open source
toolkit has enabled several other
projects~\citep[e.g.][]{kelleher2019inferring,haller2019slim,wohns2021unified,
de2021geonomics} to take advantage
of the same efficient data structures used in \msprime,
and we hope that many more will follow.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarise
some aspects that are important for simulation.

% Encoding ancestry
% JK: we can say this some other way, this just seemed like the
% most concise way. People may be irritated by being told what
% the definition of a "genome" is, but I think we need to be explicit.
Let us define a genome as the complete set of genetic material
that a child inherits from one parent. Thus, a diploid individual
has two (monoploid) genomes, one inherited from each parent. Since
each diploid individual lies at the end of two distinct lineages
of descent, they will be represented by \emph{two} places (nodes)
in any genealogical tree. In the tree sequence encoding a
\emph{node} therefore corresponds to a single genome, which
is associated with its creation time (and other optional information),
and recorded in a simple tabular format (Fig.~\ref{fig-ts-example}).
Genetic inheritance between genomes (nodes) % delete?
is defined by edges.
An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
% JK: This is potentially confusing. Changed "genomic interval"
% to "chromosome" interval to avoid talking about "the" genome.
of the contiguous chromosomal segment over which
the child genome inherited genetic material from the parent genome.
Parent and child nodes may correspond to ancestor and descendant
genomes separated by many generations.
Critically, edges can span multiple trees  along the
genome (usually referred to as ``marginal'' trees),
and identical node IDs across different trees corresponds
to the same ancestral genome.
For example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral genomes and recording the
contiguous segments over which parent-child relationships exist,
generalises the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
Note that this formulation is fully compatible with
the concept of an Ancestral Recombination Graph (ARG)
and any ARG topology can be fully and efficiently encoded
in the node and edge tables illustrated in Fig.~\ref{fig-ts-example};
see the \nameref{sec-arg} section below for
more details.

@@

The vast genomic datasets produced during the SARS-CoV-2 pandemic
have highlighted the advantages of storing genetic variation data using the underlying trees.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format (consisting of a Newick tree
and associated mutations in a binary format) and
the \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustrate the advantages of the general
approach of storing ancestry and mutations rather than sequences,
they do not generalise beyond their immediate settings, and
no software library support is available.
% \msprime, on the other hand, provides a fast, powerful, and well-tested
% generic interface for generating and storing mutations on trees.

@@

Part of the reason for this poor record of software reuse and modularity is the
lack of standardised file formats, and in particular, the absence of common
library infrastructure to abstract the details of interchanging simulation
data. Although \msprime\ also supports simulating both ancestry and mutations,
the two aspects are functionally independent within the software; both ancestry and mutation simulators are present in \msprime\
for reasons of convenience and history,
and could be split into separate packages.
The efficient C and Python interfaces for
\tskit\ make it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for operations being performed
in a different software package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real
data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

@@

\subsection*{Gene conversion}
Gene conversion is a form of recombination that results in the transfer
of a short segment of genetic material,
for example between homologous chromosomes~\citep{chen2007gene}.
Since gene conversion impacts much shorter segments than
crossover recombination (typically below 1kb) it
affects patterns of linkage disequilibrium differently~\citep{korunes2017gene}.
\cite{wiuf2000coalescent} modelled gene conversion in the coalescent
via a rate at which gene conversion events are initiated
along the genome and a geometrically distributed tract length.
In terms of the ancestral process, gene conversion differs from
crossover recombination (as described in the previous section)
in that it extracts a short tract of ancestry into
an independent lineage, rather than splitting ancestry
to the left and right of a given breakpoint.
We have implemented this model of gene
conversion in \msprime\ 1.0, and validated the output against
\ms\ and analytical results~\citep{wiuf2000coalescent}.

@@

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution,
and so we compare the performance of \msprime\ with gene conversion to
two specialised bacterial simulators,
\SimBac~\citep{brown2016simbac} and \FastSimBac~\citep{demaio2017the}.
Figure~\ref{fig-gc-perf}A shows that \msprime\ is far more efficient than
both \SimBac\ and the SMC-based approximation \FastSimBac.
Figure~\ref{fig-gc-perf}B shows that \msprime\ requires somewhat more memory
than \FastSimBac, (as expected since \FastSimBac\ uses a left-to-right
SMC approximation) but is still reasonably modest at around 1GiB for a simulation
of 500 whole \emph{E.~coli} genomes.
However, \msprime\ is currently lacking many of the specialised features
required to model bacteria, and so an important avenue for future work
is to add features such as circular genomes
and bacterial gene transfer~\citep{baumdicker2014AGTG}.

\subsection*{Demography}
\label{sec-demography}
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealised cases such as stepping-stone or
island models, or specialised cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are supported, which allow for
varying population sizes and growth rates, changing migration rates over time,
as well as population splits, admixtures and pulse migrations.

@@

A common approach to modelling the effect of demographic history on genealogies
is to assume that effective population size ($N_e$) changes in discrete steps
which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
keightley2007joint,li2011inference}. In this setting of piecewise constant
$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
bottlenecks are short both on the timescale of coalescence and mutations,
there may be little information about the duration of a bottleneck
($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
and generate a sudden burst of coalescence events (a multiple merger event) in
the genealogy. The strength of the bottleneck $B$ can be thought of as an
(imaginary) time period during which coalescence events are collapsed,
i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
single pair of lineages entering the bottleneck coalesce during the bottleneck
is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
attractive and both analytic results and empirical
inference~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
bunnefeld2015inferring} have been developed under this model, there has
been no software available to simulate data under instantaneous
bottleneck histories.

@@

Kingman's coalescent assumes that only two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation, there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16,vendrami2021sweepstake},
where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation~\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modelled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modelling the effects of selection on
gene genealogies
\citep{Gillespie909,DS04,desai2013genetic,neher2013genealogies,schweinsberg2017rigorous}.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
to represent the stochastic process of the coalescent with recombination as a graph.
This formulation is complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages. There is the ``big'' ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral
material~\citep{ethier1990two},
% and
% consequently wait an exponentially large time until all sampled lineages
% coalesce into a single ancestor~\citep{ethier1990two},
and the ``little'' ARG in which we only track genetic ancestors.
Over time, usage of the term
has shifted away from its original definition as a stochastic process,
to being interpreted as a representation of a particular genetic ancestry
as a graph, without necessarily following the specific details of the Griffiths
formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
Under the latter interpretation,
the tree sequence encoding of genetic ancestry (described above)
clearly \emph{is} an ARG: the nodes and edges
define a graph in which edges are annotated with the
set of disjoint genomic intervals
through which ancestry flows.

>>>>

\begin{abstract}
Stochastic simulation is a key tool in population genetics,
since the models involved are often analytically intractable
and simulation is usually the only way of obtaining ground-truth data to
evaluate inferences.
Because of this necessity, a large number
of specialized simulation programs have been developed, each
filling a particular niche, but with largely overlapping functionality
and a substantial duplication of effort.
Here, we introduce \msprime\ version 1.0, which
efficiently implements ancestry and mutation simulations based on
the succinct tree sequence data structure and \tskit\ library.
We summarize \msprime's many features, and show that
its performance is excellent, often many times faster
and more memory efficient than specialized alternatives.
These high-performance features have been thoroughly
tested and validated, and built using a collaborative,
open source development model,
which reduces duplication of effort and promotes
software quality via community engagement.
% "Bazaar" development model: https://en.wikipedia.org/wiki/The_Cathedral_and_the_Bazaar
\end{abstract}

@@

The coalescent
process~\citep{kingman1982coalescent,kingman1982genealogy,hudson1983testing,
tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is an efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from an idealized population model, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data,
and used these simulations to characterize an analytically
intractable distribution.
This inherent efficiency, and the great utility of simulations for a wide
range of purposes, has led to dozens of different tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

Two technological developments of recent years, however,
pose major challenges to most existing simulation methods.
Firstly, fourth-generation sequencing technologies have made
complete chromosome-level assemblies possible~\citep{miga2020telomere},
% Some citations here? Darwin tree of life maybe?
and high quality assemblies are now available for many species.
Thus, modeling genetic variation data as a series of unlinked
non-recombining loci
is no longer a reasonable approximation, and
we must fully account for recombination.
However, while a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination is far more complex,
and programs such as
Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination.
The second challenge facing simulation methods is that
sample sizes in genetic studies have grown very quickly
in recent years, enabled by the precipitous fall in genome sequencing costs.
Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm~\citep{staab2015scrm} simply cannot
cope with such a large number of samples.

@@

\section*{Results}
In the following sections we describe the main features of \msprime\ 1.0,
focusing on the aspects that are either new for this version, or in which
our approach differs significantly from classical methods
(summarized in Table~\ref{tab-v1-summary}).
Where appropriate,
we benchmark \msprime\ against other simulators, but the comparisons are
illustrative and not intended to be systematic or exhaustive. Please
see~\cite{kelleher2016efficient} for a performance comparison of
\msprime\ against simulators such as \ms, \msms, and \scrm.

\begin{table}
\small{
{ % begin box to localize effect of arraystretch change
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{lp{6.5cm}}
Interface
& Separation of ancestry and mutation simulations.
Ability to store arbitrary metadata along with simulation
results, and automatic recording of provenance information
for reproducibility.
Jupyter notebook~\citep{kluyver2016jupyter} integration.
Rich suite of analytical and visualization methods via \tskit\ library. \\

Ancestry
& SMC, SMC', Beta- and Dirac-coalescent, discrete time Wright-Fisher,
and selective sweep models. Instantaneous bottlenecks.
Discrete or continuous genomic coordinates, arbitrary ploidy,
gene conversion.
Output full ARG with recombination nodes, ARG likelihood calculations.
Record full migration history and census
events.  Improved performance for large numbers of populations.
Integration with forward simulators such as \SLiM\ and \fwdpy\
(``recapitation'').
\\

@@

The majority of simulation packages are controlled either through
a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a text-based input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or a mixture of both.
Command line interfaces make it easy to run simple
simulations, but as model complexity and the number of parameters increase,
they become difficult to understand and
error-prone~\citep{ragsdale2020lessons,gower2021demes}.
Specifying parameters through a text file alleviates this problem to a degree,
but lacks flexibility, for example, when running simulations with parameters
drawn from a distribution. In practice, for any reproducible simulation
project users will write a script
to generate the required command lines or input parameter files,
invoke the simulation engine, and process the results in some way.
This process is cumbersome and labor intensive, and
a number of packages have been developed
to allow simulations to be run directly in a high-level
scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.

The more recent trend has been to move away from this file and command-line
driven approach and to instead provide direct interfaces to the simulation
engines via an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
The primary interface for \msprime\ is through a thoroughly documented
Python
API, which has encouraged the development of an ecosystem of downstream
% Found these through the Dependent packages on GitHub
tools~\citep{terhorst2017robust,chan2018likelihood,spence2019inference,
adrion2020community,adrion2020predicting, kamm2020efficiently,
mckenzie2020ipcoal, montinaro2020revisiting,
de2021geonomics,rivera2021simulation}.
As well as providing a stable and efficient platform for building
downstream applications, \msprime's Python API makes it much easier to
build reproducible simulation pipelines, as the entire workflow can
be encapsulated in a single script, and package and version
dependencies explicitly stated using the \texttt{pip}
or \texttt{conda} package managers.
For example, the errors made
in the influential simulation analysis of
\cite{martin2017human} were only detected because the pipeline
could be easily run and reanalyzed~\citep{ragsdale2020lessons,martin2020erratum}.

\begin{figure*}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualization of the separation between ancestry and mutation
        simulation. (A) The result of an invocation of \texttt{sim\_ancestry}
        is two trees along a 1kb chunk of genome relating three diploid samples.
        Each diploid individual consists of two genomes (or nodes), indicated
        by color.
        (B) This ancestry is provided as the input to \texttt{sim\_mutations},
        which adds mutations.
        Graphics produced using \tskit's \texttt{draw\_svg} method.
    }
\end{figure*}

@@

One of the key reasons for \msprime's substantial performance advantage
over other simulators~\citep{kelleher2016efficient}
is its use of the ``succinct tree sequence''
data structure to represent simulation results.
The succinct tree sequence (usually abbreviated to ``tree sequence'')
was introduced by \cite{kelleher2016efficient}
to concisely encode genetic ancestry and sequence variation
and was originally implemented as part of \msprime.
We subsequently extracted the core tree sequence functionality
from \msprime\ to create the \tskit\ library,
which provides a large suite of tools for processing genetic ancestry
and variation data via APIs
in the Python and C languages~\citep{tskit2021tskit}.
The availability of \tskit\ as a liberally licensed (MIT) open source
toolkit has enabled several other
projects~\citep[e.g.][]{kelleher2019inferring,haller2019slim,wohns2021unified,
de2021geonomics} to take advantage
of the same efficient data structures used in \msprime,
and we hope that many more will follow.
While a full discussion of tree sequences and the capabilities
of \tskit\ is beyond the scope of this article, we summarize
some aspects that are important for simulation.

% Encoding ancestry
% JK: we can say this some other way, this just seemed like the
% most concise way. People may be irritated by being told what
% the definition of a "genome" is, but I think we need to be explicit.
Let us define a genome as the complete set of genetic material
that a child inherits from one parent. Thus, a diploid individual
has two (monoploid) genomes, one inherited from each parent. Since
each diploid individual lies at the end of two distinct lineages
of descent, they will be represented by \emph{two} places (nodes)
in any genealogical tree. In the tree sequence encoding a
\emph{node} therefore corresponds to a single genome, which
is associated with its creation time (and other optional information),
and recorded in a simple tabular format (Fig.~\ref{fig-ts-example}).
Genetic inheritance between genomes (nodes) % delete?
is defined by edges.
An \emph{edge} consists of a
parent node, a child node and the left and right coordinates
% JK: This is potentially confusing. Changed "genomic interval"
% to "chromosome" interval to avoid talking about "the" genome.
of the contiguous chromosomal segment over which
the child genome inherited genetic material from the parent genome.
Parent and child nodes may correspond to ancestor and descendant
genomes separated by many generations.
Critically, edges can span multiple trees  along the
genome (usually referred to as ``marginal'' trees),
and identical node IDs across different trees corresponds
to the same ancestral genome.
For example, in Fig.~\ref{fig-ts-example} the branch from node
0 to 4 is present in both marginal trees, and represented
by a single edge (the first row in the edge table).
This simple device, of explicitly associating tree nodes
with specific ancestral genomes and recording the
contiguous segments over which parent-child relationships exist,
generalizes the original
``coalescence records'' concept~\citep{kelleher2016efficient},
and is the key to the efficiency of tree
sequences~\citep{
kelleher2018efficient,kelleher2019inferring,ralph2020efficiently}.
Note that this formulation is fully compatible with
the concept of an Ancestral Recombination Graph (ARG)
and any ARG topology can be fully and efficiently encoded
in the node and edge tables illustrated in Fig.~\ref{fig-ts-example};
see the \nameref{sec-arg} section below for
more details.

@@

The vast genomic datasets produced during the SARS-CoV-2 pandemic
have highlighted the advantages of storing genetic variation data using the underlying trees.
\cite{turakhia2021ultrafast} propose the
Mutation Annotated Tree (MAT)
format (consisting of a Newick tree
and associated mutations in a binary format) and
the \texttt{matUtils} program as an efficient way to store
and process large viral datasets~\citep{mcbroome2021daily},
achieving excellent compression and processing performance.
Similarly, \phastsim~\citep{demaio2021phastsim}
was developed to simulate sequence evolution on such large SARS-CoV-2
phylogenies, and also outputs a Newick tree annotated with mutations (not
in MAT format) to avoid the bottleneck of generating
and storing the simulated sequences.
While these methods illustrate the advantages of the general
approach of storing ancestry and mutations rather than sequences,
they do not generalize beyond their immediate settings, and
no software library support is available.
% \msprime, on the other hand, provides a fast, powerful, and well-tested
% generic interface for generating and storing mutations on trees.

@@

Part of the reason for this poor record of software reuse and modularity is the
lack of standardized file formats, and in particular, the absence of common
library infrastructure to abstract the details of interchanging simulation
data. Although \msprime\ also supports simulating both ancestry and mutations,
the two aspects are functionally independent within the software; both ancestry and mutation simulators are present in \msprime\
for reasons of convenience and history,
and could be split into separate packages.
The efficient C and Python interfaces for
\tskit\ make it straightforward to add further information to
an existing file, and because of its efficient data interchange mechanisms,
there is no performance penalty for operations being performed
in a different software package.
Thanks to this interoperability, \msprime's
mutation generator can work with \emph{any} \tskit\ tree sequence,
be it simulated using \SLiM~\citep{haller2019slim} or
\fwdpy~\citep{thornton2014cpp}, or estimated from real
data~\citep{kelleher2019inferring,speidel2019method,wohns2021unified}.
It is a modular
component intended to fit into a larger software ecosystem, and
is in no way dependent on \msprime's ancestry simulator.

@@

\subsection*{Gene conversion}
Gene conversion is a form of recombination that results in the transfer
of a short segment of genetic material,
for example between homologous chromosomes~\citep{chen2007gene}.
Since gene conversion impacts much shorter segments than
crossover recombination (typically below 1kb) it
affects patterns of linkage disequilibrium differently~\citep{korunes2017gene}.
\cite{wiuf2000coalescent} modeled gene conversion in the coalescent
via a rate at which gene conversion events are initiated
along the genome and a geometrically distributed tract length.
In terms of the ancestral process, gene conversion differs from
crossover recombination (as described in the previous section)
in that it extracts a short tract of ancestry into
an independent lineage, rather than splitting ancestry
to the left and right of a given breakpoint.
We have implemented this model of gene
conversion in \msprime\ 1.0, and validated the output against
\ms\ and analytical results~\citep{wiuf2000coalescent}.

@@

Gene conversion is particularly useful to model homologous recombination in
bacterial evolution,
and so we compare the performance of \msprime\ with gene conversion to
two specialized bacterial simulators,
\SimBac~\citep{brown2016simbac} and \FastSimBac~\citep{demaio2017the}.
Figure~\ref{fig-gc-perf}A shows that \msprime\ is far more efficient than
both \SimBac\ and the SMC-based approximation \FastSimBac.
Figure~\ref{fig-gc-perf}B shows that \msprime\ requires somewhat more memory
than \FastSimBac, (as expected since \FastSimBac\ uses a left-to-right
SMC approximation) but is still reasonably modest at around 1GiB for a simulation
of 500 whole \emph{E.~coli} genomes.
However, \msprime\ is currently lacking many of the specialized features
required to model bacteria, and so an important avenue for future work
is to add features such as circular genomes
and bacterial gene transfer~\citep{baumdicker2014AGTG}.

\subsection*{Demography}
\label{sec-demography}
One of the key applications of population genetic simulations is to generate
data for complex demographies. Beyond idealized cases such as stepping-stone or
island models, or specialized cases such as isolation-with-migration models,
analytical results are rarely possible. Simulation is therefore integral to the
development and evaluation of methods for demographic inference. The demography
model in \msprime\ is directly derived from the approach used in \ms, and
supports an arbitrary number of randomly mating populations exchanging migrants
at specified rates. A range of demographic events are supported, which allow for
varying population sizes and growth rates, changing migration rates over time,
as well as population splits, admixtures and pulse migrations.

@@

A common approach to modeling the effect of demographic history on genealogies
is to assume that effective population size ($N_e$) changes in discrete steps
which define a series of epochs~\citep{griffiths1994sampling, marth2004allele,
keightley2007joint,li2011inference}. In this setting of piecewise constant
$N_e$, capturing a population bottleneck requires three epochs: $N_e$ is
reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and
recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If
bottlenecks are short both on the timescale of coalescence and mutations,
there may be little information about the duration of a bottleneck
($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to
assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$)
and generate a sudden burst of coalescence events (a multiple merger event) in
the genealogy. The strength of the bottleneck $B$ can be thought of as an
(imaginary) time period during which coalescence events are collapsed,
i.e.\ there is no growth in genealogical branches during $B$ and the probability that a
single pair of lineages entering the bottleneck coalesce during the bottleneck
is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is
attractive and both analytic results and empirical
inference~\citep{griffiths1994sampling, birkner2009modified,galtier2000detecting,
bunnefeld2015inferring} have been developed under this model, there has
been no software available to simulate data under instantaneous
bottleneck histories.

@@

Kingman's coalescent assumes that only two ancestral lineages can merge at
each merger event. Although this is generally a reasonable approximation, there
are certain situations in which the underlying mathematical assumptions are
violated. For example in certain highly fecund organisms
\citep{hedgecock_94,B94,HP11,A04,irwin16,vendrami2021sweepstake},
where individuals have the ability
to produce numbers of offspring on the order of the population size and
therefore a few individuals may produce the bulk of the offspring in any given
generation~\citep{hedgecock_94}. These population dynamics violate basic
assumptions of the Kingman coalescent, and are better modeled by
`multiple-merger' coalescents \citep{DK99,P99,S99,S00,MS01}, in which more than
two lineages can merge in a given event. Multiple-merger coalescent processes
have also been shown to be relevant for modeling the effects of selection on
gene genealogies
\citep{Gillespie909,DS04,desai2013genetic,neher2013genealogies,schweinsberg2017rigorous}.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
to represent the stochastic process of the coalescent with recombination as a graph.
This formulation is complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realization of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages. There is the ``big'' ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral
material~\citep{ethier1990two},
% and
% consequently wait an exponentially large time until all sampled lineages
% coalesce into a single ancestor~\citep{ethier1990two},
and the ``little'' ARG in which we only track genetic ancestors.
Over time, usage of the term
has shifted away from its original definition as a stochastic process,
to being interpreted as a representation of a particular genetic ancestry
as a graph, without necessarily following the specific details of the Griffiths
formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
Under the latter interpretation,
the tree sequence encoding of genetic ancestry (described above)
clearly \emph{is} an ARG: the nodes and edges
define a graph in which edges are annotated with the
set of disjoint genomic intervals
through which ancestry flows.
